Concrete Semantics for Pushdown Analysis:\\The Essence of Summarization

### 2013-05-09 paper.tex
Removed rundandant sentences and changed cfa2 model to fix fake rebinding at function call, not return.

\abstract{GAH}

@@

Programs in higher-order languages heavily use function calls and
method dispatch as part of their control flow. Until recently, flow
analyses could not handle return flow precisely
\citep{ianjohnson:vardoulakis-lmcs11, ianjohnson:earl2010pdcfa}, which
leads to several spurious paths (and thus false positives) due to
their pervasiveness. These works, called CFA2 and PDCFA respectively,
use pushdown automata as their approximation's target model of
computation. They are hence called ``pushdown analyses.''\footnote{We
  will refer to the classic finite model analyses as ``regular
  analyses'' after regular languages.} The precision and performance
gains of these analyses are so beneficial that we believe new analyses
going forward should be using pushdown techniques. CFA2 and PDCFA have
difficult details to easily apply to an off-the-shelf semantics ---
especially if they feature non-local control transfer that breaks the
pushdown model.

...

@@

This paper lays out a common framework to talk about pushdown analysis
for higher-order languages. In particular, we extract the ``essence''
of the existing techniques into a quality of the machine semantics
that specify a language's meaning, much in the same way as
~\citet{dvanhorn:VanHorn2012Systematic} (a method we will refer to as
AAM: abstracting abstract machines). Once the machine semantics is in
this form, simple pointwise abstraction leads to the summarization
algorithms that we see in the literature. In effect, we give a
\emph{concrete} semantics to pushdown analysis.

A downside of the work providing \rackett{call/cc} is that it was an
algorithmic change to the already complicated CFA2 --- there was no
recipe for how to do this for one's favorite control operator. This
paper seeks to do just that with an operational view of what
summarization is, in essence. In other words, we give a concrete
semantics to the tricks that the analyses in the literature use in the
abstract. We also give intuitive analogies to well-established
ideas/techniques so that you too can write a pushdown analysis for
your language. In order to demonstrate the applicability of this
viewpoint, we show a new analysis for a language with composable
control. All of the semantics modeled in this paper are implemented in
full detail in PLT Redex and available at

@@

@@

PDCFA does not have some orthogonal semantic components that CFA2
features to improve precision, so we start by deriving PDCFA from an
operational semantics for the untyped lambda calculus (figure
\ref{fig:base-semantics}). The semantic spaces for the machine follow.

@@

Since PDCFA is a simpler analysis than CFA2, it enjoys a polynomial
complexity in the monovariant case. Its final ${\mathcal O}(n^6)$ form
is a naive ``step all states'' widening, which yields poor
results. The common widening uses a worklist to only step states that
need stepping --- the fact that the store is shared amongst all states
makes this a semantic difference, and not just an implementation
difference. Doing this bumps the complexity up to ${\mathcal
  O}(n^9)$\footnote{The available \emph{implementations} of PDCFA
  indeed do this and are themselves ${\mathcal O}(n^9)$.}. These
bounds are (rough) absolute worst case that is highly unlikely, if
ever, happen in practice. The worklist abstraction restricts value
flow to fewer spurious paths, often leading to a faster analysis than
a typical 0CFA.

@@

CFA2 is the first published analysis of a higher-order programming
language that could properly match calls and returns. We will show
that it fits well into the same presentation we gave for
PDCFA. Vardoulakis and Shivers had a clear goal of harnessing the
extra information a pushdown model provides to produce a
high-precision analysis that works well in practice. This resulted in
more than just the call/return matching of the previous section, which
is why we are showing the two separately. There are two orthogonal
features of the semantics:
\begin{enumerate}
\item{stack allocation for some bindings}
\item{strong updates on stack frames for resolved nondeterminism}
\end{enumerate}
The first of these is an addition to the stackless context. There is a
conservative pre-analysis that checks locally whether a binding will
never escape, and classifies references as able to use the stack frame
or not. Their criteria for a binding never escaping is that it is
never referenced in a function that is not its binder. This can be
extended in a language with more linguistic features; see Kranz's
thesis about register-allocatable bindings in the Orbit Scheme
compiler \citep{ianjohnson:kranz:thesis:1988}. A stackable reference
is one that appears in the body of the binding function, by which we
mean not within the body of nested function. They use the information
that a binding never escapes to not bother allocating it in the
heap. This has the advantage of not changing the heap, and thus leads
to less propagation. The addition of these stack frames makes the
analysis exponential in theory, though in practice they have been
observed to decrease running time in most cases.

The second of these is to ameliorate a problem they call ``fake
rebinding.'' That is, since bindings in the abstract represent several
values, we don't want to reference a variable \texttt{x} in two
different places and have it resolve to two different values. In AAM,
a variable reference non-deterministically steps to all possible
values associated with that variable. Here we want to say that once
\texttt{x} is considered to stand for value $v$, then all subsequent
references of \texttt{x} should be $v$. If they aren't, it looks as if
\texttt{x} was rebound; it hasn't, and thus it is a ``fake
rebinding.'' CFA2 does not step to all values on variable reference,
but instead carries all its values around in superposition until they
need to be observed at, say, a function call. We give a simplified
semantics that is more along the AAM style, but CFA2's approach can
easily be recovered from it\footnote{Our Redex model implements the fake rebinding fix after function return}.

@@

The widening we applied to PDCFA also applies here in order to get
CFA2. Both PDCFA and CFA2 use $\epsilon$-edges and summaries to
seemingly skip work when new stacks reach a function, but the summary
is not likely to be there, so we just drop that whole bit of
complexity from the semantics. CFA2 also included what they called
``transitive summaries'' to deal with tail calls. We insert an $\rt$
frame at every function call.  This view contends with tail calls, but
we can identify tail calls easily --- any call with an $\rt$ as its
continuation is a tail call. Tail calls are important in language
implementations for space complexity reasons
\citep{ianjohnson:clinger:tail-calls:1998}, but in an analysis, these
concerns are less important. The repeated popping of $\rt$ inserted
by what otherwise were tail-calls is synonymous with CFA2's transitive
summaries.

@@

\begin{figure}
  \centering
% Shift
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
    $\tpl{(\sshift{\mvar}{\mexpr}, \menv), \mstore, \mkont, \mmkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\vcomp{\mkont'}},\kmt,\mmkont,\mktab',\mmemo}$
    \\
    where & $\maddr = \alloc(\varsigma)$, $(\mkont',\mktab') = \approximate(\mkont, \mktab, \maddr)$
\\
% Return
   $\tpl{\mval, \mstore, \krt{\mctx}, \mmkont, \mktab, \mmemo}$
   &
   $\tpl{\mval, \mstore, \mkont, \mmkont, \mktab, \mmemo'}$
   if $\mkont \in \returns(\mktab, \mctx)$
   \\
   where & $\mmemo' = \memoize(\mmemo, \mktab, \mctx, \mval)$
  \end{tabular}
  \caption{Fixed table-based semantics for shift/reset}
  \label{fig:shift-reset-table1}
\end{figure}

>>>>

\abstract{Lorem Ipsum}

@@

Programs in higher-order languages heavily use function calls and
method dispatch as part of their control flow. Until recently, flow
analyses for higher-order languages could not handle return flow precisely
\citep{ianjohnson:vardoulakis-lmcs11, ianjohnson:earl2010pdcfa}, which
leads to several spurious paths (and thus false positives) due to
their pervasiveness. These works, called CFA2 and PDCFA respectively,
use pushdown automata as their approximation's target model of
computation. They are hence called ``pushdown analyses.''\footnote{We
  will refer to the classic finite model analyses as ``regular
  analyses'' after regular languages.} The precision and performance
gains of these analyses are so beneficial that we believe new analyses
going forward should be using pushdown techniques. CFA2 and PDCFA have
difficult details to easily apply to an off-the-shelf semantics ---
especially if they feature non-local control transfer that breaks the
pushdown model. \\

This paper gives a common, simple framework to discuss both CFA2 and
PDCFA along with a recipe to apply pushdown analysis techniques to an
off-the-shelf semantics in a manner similar to
~\citet{dvanhorn:VanHorn2010Abstracting} (a technique called
abstracting abstract machines, or AAM). We then extend the state of
the art by giving a novel analysis for delimited, composable control. \\

Testing new ideas in analysis for improving precision or increasing
performance can be a difficult venture. We contend that the machinery
that we employ in the abstract should have a concrete counterpart that
maintains the meaning of the language so that we can see their effect
without introducing approximations which might mask correctness
issues. Abstraction should be a simple process that is ``obviously
correct.'' The framework we present in this paper is applicable in the
concrete such that a pointwise abstraction leads to the pushdown
analyses in the literature.

@@

A downside of the work providing \rackett{call/cc} is that it was an
algorithmic change to the already complicated CFA2 --- there was no
recipe for how to do this for one's favorite control operator. This
paper seeks to do just that with an operational view of what
summarization is, in essence. In other words, we give a concrete
semantics to the tricks that the analyses in the literature use in the
abstract, and maintain the original meaning of the language. We also
give intuitive analogies to well-established ideas/techniques so that
you too can write a pushdown analysis for your language. In order to
demonstrate the applicability of this viewpoint, we show a new
analysis for a language with composable control. All of the semantics
modeled in this paper are implemented in full detail in PLT Redex and
available at

@@

There are common underpinnings of PDCFA and CFA2 that can be embodied
as concrete machinery in the programming language semantics: breaking
the recursive structure of continuations by indirecting them through a
table with appropriately precise keys, and memoization.  Once the
machine semantics is in this form, simple pointwise abstraction leads
to the summarization algorithms that we see in the literature. In
effect, we give a \emph{concrete} semantics to pushdown analysis.

@@

PDCFA does not have some orthogonal semantic components that CFA2
features to improve precision and is thus the simpler of the two. We
start by deriving PDCFA from an operational semantics for the untyped
lambda calculus (figure \ref{fig:base-semantics}). The semantic spaces
for the machine follow.

@@

Since PDCFA is a simpler analysis than CFA2, it enjoys a polynomial
complexity in the monovariant case. Its final ${\mathcal O}(n^6)$ form
is a naive ``step all states'' widening, which yields poor
results. The common widening uses a worklist to only step states that
need stepping --- the fact that the store is shared amongst all states
makes this a semantic difference, and not just an implementation
difference. Doing this bumps the complexity up to ${\mathcal
  O}(n^9)$. The available \emph{implementations} of PDCFA indeed do
this and are themselves ${\mathcal O}(n^9)$. These bounds are (rough)
absolute worst case that is highly unlikely, if ever, happen in
practice. The worklist abstraction restricts value flow to fewer
spurious paths, often leading to a faster analysis than a typical
${\mathcal O}(n^3)$ 0CFA.

@@

CFA2 is the first published analysis of a higher-order programming
language that could properly match calls and returns. We will show
that it fits well into the same presentation we gave for
PDCFA. Vardoulakis and Shivers had a clear goal of harnessing the
extra information a pushdown model provides to produce a
high-precision analysis that works well in practice. This resulted in
more than just the call/return matching of the previous section, which
is why we are showing the two separately. There are two orthogonal
features of the semantics:
\begin{enumerate}
\item{stack allocation for some bindings}
\item{strong updates on stack frames for resolved nondeterminism}
\end{enumerate}
The first of these is an addition to the stackless context. There is a
conservative pre-analysis that checks locally whether a binding will
never escape, and classifies references as able to use the stack frame
or not. Their criteria for a binding never escaping is that it is
never referenced in a function that is not its binder. This can be
extended in a language with more linguistic features; see Kranz's
thesis about register-allocatable bindings in the Orbit Scheme
compiler \citep{ianjohnson:kranz:thesis:1988}. A stackable reference
is one that appears in the body of the binding function, by which we
mean not within the body of a nested function. They use the information
that a binding never escapes to not bother allocating it in the
heap. This has the advantage of not changing the heap, and thus leads
to less propagation. The addition of these stack frames makes the
analysis exponential in theory, though in practice they have been
observed to decrease running time in most cases.

The second of these is to ameliorate a problem they call ``fake
rebinding.'' That is, since bindings in the abstract represent several
values, we don't want to reference a variable \texttt{x} in two
different places and have it resolve to two different values. In AAM,
a variable reference non-deterministically steps to all possible
values associated with that variable. Here we want to say that once
\texttt{x} is considered to stand for value $v$, then all subsequent
references of \texttt{x} should be $v$. If they aren't, it looks as if
\texttt{x} was rebound; it hasn't, and thus it is a ``fake
rebinding.'' CFA2 does not step to all values on variable reference,
but instead carries all its values around in superposition until they
need to be observed at, say, a function call. We give a simplified
semantics that is more along the AAM style, but CFA2's approach can
easily be recovered from it\footnote{Our Redex model implements the fake rebinding strategy CFA2 itself employs}.

@@

CFA2 also includes what they called ``transitive summaries'' to deal
with tail calls. We insert an $\rt$ frame at every function call.
This view contends with tail calls, but we can identify tail calls
easily --- any call with an $\rt$ as its continuation is a tail
call. Tail calls are important in language implementations for space
complexity reasons \citep{ianjohnson:clinger:tail-calls:1998}, but in
an analysis, these concerns are less important. The repeated popping
of $\rt$ inserted by what otherwise were tail-calls is synonymous with
CFA2's transitive summaries.

@@

\begin{figure}
  \centering
% Shift
  $\mstate \stepto \mstate' \text{ where } \maddr = \alloc(\mstate)$ \\
  \begin{tabular}{r|l}
    \hline
    $\tpl{(\sshift{\mvar}{\mexpr}, \menv), \mstore, \mkont, \mmkont, \mktab, \mmemo}$
    &
    $\tpl{(\mexpr, \extm{\menv}{\mvar}{\maddr}), \joinone{\mstore}{\maddr}{\vcomp{\mkont'}},\kmt,\mmkont,\mktab',\mmemo}$
    \\
    where & $(\mkont',\mktab') = \approximate(\mkont, \mktab, \maddr)$
\\
% Return
   $\tpl{\mval, \mstore, \krt{\mctx}, \mmkont, \mktab, \mmemo}$
   &
   $\tpl{\mval, \mstore, \mkont, \mmkont, \mktab, \mmemo'}$
   if $\mkont \in \returns(\mktab, \mctx)$
   \\
   where & $\mmemo' = \memoize(\mmemo, \mktab, \mctx, \mval)$
  \end{tabular}
  \caption{Fixed table-based semantics for shift/reset}
  \label{fig:shift-reset-table1}
\end{figure}
